Wallace Session 2023-06-07
Erica Johnson
Cloud Forest
================

Please find below the R code history from your *Wallace* v1.9.0 session.

### Package installation

Wallace uses the following R packages that must be installed and loaded
before starting.

```{r}
library(spocc)
library(spThin)
library(dismo)
library(rgeos)
library(raster)
library(ENMeval)
library(wallace)
```

------------------------------------------------------------------------

## Analysis for *Cloud forest* (Cf)

User CSV path with occurrence data. 

```{r}
# NOTE: provide the folder path of the .csv file
occs_path <- "/home/erica/PhD/ENMRandomization/ENMReal/data/MexVeg/OccData/RandFilt"
occs_path <- file.path(occs_path, "Cloud_forest25.csv")
# get a list of species occurrence data
userOccs_Cf <- occs_userOccs(
  txtPath = occs_path, 
  txtName = "Cloud_forest25", 
  txtSep = ",", 
  txtDec = ".")
occs_Cf <- userOccs_Cf$Cloud_forest$cleaned
```

### Obtain environmental data

Using user-specified variables.

```{r}
## Specify the directory with the environmental variables
dir_envs_Cf <- "/home/erica/PhD/ENMRandomization/ENMReal/data/MexVeg/Env"
envs_path <- file.path(dir_envs_Cf, dir(dir_envs_Cf, pattern = ".grd"))

# All preds: Chelsa bioclim + cloud cover + soil
envs_Cf <- envs_userEnvs(
  rasPath = envs_path,
  rasName = dir(dir_envs_Cf, pattern = ".grd$"),
  doBrick = FALSE)
```

```{r}
occs_xy_Cf <- occs_Cf[c('longitude', 'latitude')]
occs_vals_Cf <- as.data.frame(raster::extract(envs_Cf, occs_xy_Cf))
# remove occurrence records with NA environmental values
occs_Cf <- occs_Cf[!(rowSums(is.na(occs_vals_Cf)) > 1), ]
# also remove variable value rows with NA environmental values
occs_vals_Cf <- na.omit(occs_vals_Cf)
# add columns for env variable values for each occurrence record
occs_Cf <- cbind(occs_Cf, occs_vals_Cf)
```

### Process environmental data

Sampling the entire available environmental background using a “point buffers” method with a 0.25 degree buffer.
```{r}
# Generate background extent 
bgExt_Cf <- penvs_bgExtent(
  occs = occs_Cf,
  bgSel = "point buffers",
  bgBuf = 0.25)
# Mask environmental data to provided extent
bgMask_Cf <- penvs_bgMask(
  occs = occs_Cf,
  envs = envs_Cf,
  bgExt = bgExt_Cf)
# Sample the full environmental background from the provided area
bgSample_Cf <- penvs_bgSample(
  occs = occs_Cf,
  bgMask =  bgMask_Cf,
  bgPtsNum = sum(!is.na(getValues(bgMask_Cf[[1]])))) #ADDED BY EEJ 

# Extract values of environmental layers for each background point
bgEnvsVals_Cf <- as.data.frame(raster::extract(bgMask_Cf,  bgSample_Cf))


##Add extracted values to background points table
bgEnvsVals_Cf <- cbind(scientific_name = paste0("bg_", "Cloud forest"), bgSample_Cf,
                       occID = NA, year = NA, institution_code = NA, country = NA,
                            state_province = NA, locality = NA, elevation = NA,
                            record_type = NA, bgEnvsVals_Cf)


#Remove rows with NAs to ensure all predictor sets have equal number of bg
bgEnvsVals_Cf<- bgEnvsVals_Cf %>% tidyr::drop_na(any_of(names(envs_Cf)))

#Remove those coords in bgSample
bgSample_Cf <- dplyr::inner_join(bgSample_Cf, 
                    bgEnvsVals_Cf, by = c('longitude', 'latitude'))
```

```{r}
#Exporting environmental background rasters
bg.dir <- "/home/erica/PhD/ENMRandomization/ENMReal/data/MexVeg/Bg"
writeRaster(bgMask_Cf, paste(bg.dir, "CF_Bg.grd", sep = "/"), format = "raster", overwrite= T)
```

### Partition occurrence data

Partition occurrences and background points for model training and
validation using “spatial block”, a spatial partition method with an
aggregation factor of 2.

```{r}
# R code to get partitioned data
groups_Cf <- part_partitionOccs(
  occs = occs_Cf ,
  bg =  bgSample_Cf, 
  method = "cb2",
  bgMask = bgMask_Cf,
  aggFact = 2) 
```


### Build and Evaluate Niche Model

Generating a species distribution model using the maxnet algorithm
as implemented in ENMeval V2.0 (with clamping = TRUE). For tuning using
L, LQ, H, LQH, P, T feature classes and regularization multipliers in the 1, 5
range increasing by 0.5. Using a total of 1 categorical predictor
variables.

```{r}
# Run maxent model for the selected species for each predictor set

#Bioclim only
model_Cf_bioclim<- model_maxent(
  occs = occs_Cf,
  bg = bgEnvsVals_Cf[1:30],
  user.grp = groups_Cf, 
  bgMsk = bgMask_Cf[[1:19]],
  rms = c(1,5), 
  rmsStep = 0.5,
  fcs = c('L','LQ','H','LQH','LQHP'),
  clampSel = TRUE,
  algMaxent = "maxent.jar",
  parallel = TRUE,
  numCores = 10)

#Bioclim + cloud
model_Cf_biocloud <- model_maxent(
  occs = occs_Cf,
  bg = bgEnvsVals_Cf[1:34],
  user.grp = groups_Cf, 
  bgMsk = bgMask_Cf[[1:23]],
  rms = c(1,5), 
  rmsStep = 0.5,
  fcs = c('L','LQ','H','LQH','LQHP'),
  clampSel = TRUE,
  algMaxent = "maxent.jar",
  parallel = TRUE,
  numCores = 10)

#Cloud only
cloudEnvs = c(1:11, 31:34)
model_Cf_cloud <- model_maxent(
  occs = occs_Cf,
  bg = bgEnvsVals_Cf[cloudEnvs],
  user.grp = groups_Cf, 
  bgMsk = bgMask_Cf[[20:23]],
  rms = c(1,5), 
  rmsStep = 0.5,
  fcs = c('L','LQ','H','LQH','LQHP'),
  clampSel = TRUE,
  algMaxent = "maxent.jar",
  parallel = TRUE,
  numCores = 10)
```

```{r}
#exporting ENMeval results
dir_eval <- "/home/erica/PhD/ENMRandomization/ENMReal/outputs/MexVeg/ENMeval"
dir_mod <- "/home/erica/PhD/ENMRandomization/ENMReal/outputs/MexVeg/Models"

#Bioclim only
write.csv(model_Cf_bioclim@results, paste(dir_eval, "CF_Bioclim.csv", sep = "/"), 
          row.names = FALSE)
write.csv(model_Cf_bioclim@results.partitions, paste(dir_eval, "partitions",
                                             "CF_Bioclim_part.csv", sep = "/"), row.names = FALSE)
saveRDS(model_Cf_bioclim, paste(dir_mod, "CF_Bioclim.rds", sep = "/"))

#Bioclim + cloud
write.csv(model_Cf_biocloud@results, paste(dir_eval, "CF_BioCloud.csv", sep = "/"), 
          row.names = FALSE)
write.csv(model_Cf_biocloud@results.partitions, paste(dir_eval, "partitions",
                                             "CF_BioCloud_part.csv", sep = "/"), row.names = FALSE)
saveRDS(model_Cf_biocloud, paste(dir_mod, "CF_BioCloud.rds", sep = "/"))


#Cloud only
write.csv(model_Cf_cloud@results, paste(dir_eval, "CF_Cloud.csv", sep = "/"), 
          row.names = FALSE)
write.csv(model_Cf_cloud@results.partitions, paste(dir_eval, "partitions",
                                             "CF_Cloud_part.csv", sep = "/"), row.names = FALSE)
saveRDS(model_Cf_cloud, paste(dir_mod, "CF_Cloud.rds", sep = "/"))
```

### Select optimal model 

```{r}
#' Selecting optimal models by lowest AICc
#' @param sp character string of species name
#' @param eval.dir Directory where ENMevaluation result tables are be saved
#' @param mod.dir Directory where ENMevaluation objects are saved 
#' @param preds character indicating the file name to match the predictor variables with

model_select <- function(sp, eval.dir, mod.dir, preds){
  library(dplyr)
  #load ENMeval table and sorting by AICc
  tbl <- read.csv(file.path(eval.dir, dir(eval.dir, pattern = paste(sp, preds, sep = "_"))))
  tbl <- as_tibble(tbl)%>%
      #Removing NAs
      dplyr::filter(!is.na(AICc)) %>%
      #Selecting co-optimal models by lowest AICc
      dplyr::filter(AICc == min(AICc))%>%
      dplyr::mutate(scientific_name = rep(sp, nrow(.))) %>% 
      dplyr::mutate(predictors = rep(preds, nrow(.)))
    
    args <- list(fc = tbl$fc, rm = tbl$rm)
  #exporting tuning arguments of co-optimal models
  saveRDS(args, paste(mod.dir,paste(sp, preds, "args.rds", sep = "_"), sep = "/"))
  return(tbl)
}
  
```


```{r}
preds <- c("Bioclim", "BioCloud", "Cloud")
mCf_select <- lapply(preds, function(p){
  mod <- model_select(sp = "CF", eval.dir = dir_eval, mod.dir = dir_mod, preds = p)
}) 


mCf_select <- dplyr::bind_rows(mCf_select) %>% 
  dplyr::relocate(scientific_name, predictors)

if(file.exists(paste(dir_mod, "ENM_OptArgs.csv", sep = "/")) == F){
    write.csv(mCf_select, paste(dir_mod, "ENM_OptArgs.csv", sep = "/"), 
              row.names =  F)
  } else {
    write.table(mCf_select, paste(dir_mod, "ENM_OptArgs.csv", sep = "/"),
                append = T, row.names = F, col.names = F, sep = ",")
    
  }
```

### Visualize

Generate a raster of the Maxent generated model with no threshold (cloglog).

```{r}
# Select current model and obtain raster prediction

predSel_Cf <- lapply(preds, function(p){
  args <- mCf_select %>% dplyr:: filter(predictors == p)%>% 
    dplyr::select(tune.args)
  args <- args$tune.args
  
  if(p == "Bioclim"){
    m_Cf <- model_Cf_bioclim@models[[args]]
    bgMask <- bgMask_Cf[[1:19]]
  } else if (p == "BioCloud"){
    m_Cf <- model_Cf_biocloud@models[[args]]
    bgMask <- bgMask_Cf[[1:23]]
  } else if (p == "Cloud"){
    m_Cf <- model_Cf_cloud@models[[args]]
    bgMask <- bgMask_Cf[[20:23]]
  } 
  predSel_Cf <- dismo::predict(m_Cf, bgMask, type = "cloglog", clamp = TRUE)
  
  #export prediction raster
  writeRaster(predSel_Cf,
              paste("/home/erica/PhD/ENMRandomization/ENMReal/outputs/MexVeg/Rasters",
              paste0("CF_",p,".tif"), sep = "/"),
              format = "GTiff", overwrite = T)
})
```

Obtaining and exporting lambdas files
```{r}
curves.dir <- "/home/erica/PhD/ENMRandomization/ENMReal/outputs/MexVeg/ResponseCurves/"

lapply(preds[2:3], function(p){
  args <- mCf_select %>% dplyr:: filter(predictors == p)%>% 
    dplyr::select(tune.args)
  args <- args$tune.args
  if(p == "Bioclim"){
    m_Cf <- model_Cf_bioclim@models[[args]]
  } else if (p == "BioCloud"){
    m_Cf <- model_Cf_biocloud@models[[args]]
  } else if (p == "Cloud"){
    m_Cf <- model_Cf_cloud@models[[args]]
  } else if (p == "AllPreds"){
    m_Cf <- model_Cf_soil@models[[args]]
  }
  lambdas <- lambdasDF(m_Cf, alg = "maxent.jar")
  write.csv(lambdas, paste(curves.dir, paste0("CF_", p, "_lambdas.csv"), sep = "/"))
  
#pdf(file = "/home/erica/PhD/ENMRandomization/ENMReal/outputs/MexVeg/ResponseCurves/CF_Cloud_respcurve.pdf")
#lapply(names(bgMask_Cf), maxnet::response.plot, x = m_Cf, var = bgMask, type = "cloglog")
#while(!is.null(dev.list())) dev.off()
})
```


## Calculating model thresholds

Calculating the minimum training presence and 10 pct threshold. 

```{r}
#ADDED TO WALLACE RMD BY EEJ

thresh <- lapply(predSel_Cf, function(p){
pred_vals_Cf <- raster::extract(p, occs_xy_Cf)

MTP <- min(na.omit(pred_vals_Cf))
pct10 <-  rev(sort(pred_vals_Cf))[(ceiling(length(pred_vals_Cf) * 0.9))] 
predictors <- names(p)
th <- data.frame(predictors, MTP, pct10)

if(file.exists(paste(dir_mod, "ENM_threshold.csv", sep = "/")) == F){
    write.csv(th, paste(dir_mod, "ENM_threshold.csv", sep = "/"), 
              row.names =  F)
  } else {
    write.table(th, paste(dir_mod, "ENM_threshold.csv", sep = "/"),
                append = T, row.names = F, col.names = F, sep = ",")
}
})

```

## Project model

Projecting the model to the Carso Huasteco + Gran Sierra Plegada

```{r}
# Loading the Carso Huasteco shapefile and defining it as the projection area
proj_path <- "/home/erica/PhD/ENMRandomization/ENMReal/data/MexVeg/CountryData"
proj_userExt_Cf <- rgdal::readOGR(file.path(proj_path, "SMOr.shp"), 
                                  layer = "SMOr")

# Generate a projection of the model to the desired area
proj_area_Cf <- lapply(preds[2:3], function(p){
  
  args <- mCf_select %>% dplyr:: filter(predictors == p)%>% 
    dplyr::select(tune.args)
  args <- args$tune.args
  
  if(p == "Bioclim"){
    m_Cf <- model_Cf_bioclim
    # Create object of projection variables
    projAreaEnvs_Cf <- envs_Cf[[1:19]]
  } else if (p == "BioCloud"){
    m_Cf <- model_Cf_biocloud
    projAreaEnvs_Cf <- envs_Cf[[1:23]]
  } else if (p == "Cloud"){
    m_Cf <- model_Cf_cloud
    projAreaEnvs_Cf <- envs_Cf[[20:23]]
  }
  
  proj_area_Cf <- proj_area(
    evalOut = m_Cf,
    curModel = args,
    envs = projAreaEnvs_Cf , 
    outputType = "cloglog",
    alg = "maxent.jar",
    clamp = TRUE,
    pjExt = proj_userExt_Cf) 
  
  out_path <- "/home/erica/PhD/ENMRandomization/ENMReal/outputs/MexVeg/Rasters"
  
  writeRaster(proj_area_Cf$projArea, 
            paste(out_path, paste0("CF_", p,"_pj.tif"), sep = "/"), format = "GTiff", 
            overwrite= T)
})
```
